---
title: "Postelsia + Heatwave Data Exploration"
format:
  html:
    code-fold: true
    code-summary: "Show Code"  # Optional: Custom label for the folded code
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = "/Volumes/GERRATY/Postelsia_Collapse")
```

In this document, we begin some preliminary investigation of datasets that we can leverage to explore the impact of the 2014-2016 marine heatwave on *Postelsia palmaeformis.* Here, we do some data exploration + visualization to help guide project development.

There are several datasets we can draw from for the project, and perhaps we can start by discussing the benefits and drawbacks of each:

**Postelsia Data**

-   Postelsia plots (high coverage of sites/years, lots of signal)

-   Repeat photos (low coverage, can be directly linked with mussel data from the same photos)

-   Biodiversity surveys (not the best, I understand, but I don't know these datasets well)

**Mussel Data**

-   Repeat photos (low coverage, but can be directly linked with postelsia data from the same photos)

-   Biodiversity surveys (low annual coverage, need to check to be sure sites have been surveyed somewhat consistently pre-post heatwave)

-   Targeted assemblage photo quads (?) - (not intended use, I understand)

**Sea Star Data**

-   Sea star plots (pretty solid, see visualizations and notes below)

**Temperature Data**

-   From HOBO loggers (water temp) and/or weather stations (air temp). Do we need? Let's discuss.

```{r}

# Part 1: Load Packages --------------------------------------------------

# Load packages
packages <- c("tidyverse", "mgcv", "ggthemes")

pacman::p_load(packages, character.only = TRUE); rm(packages)

setwd(here::here())
```

------------------------------------------------------------------------

#### Dataset 1: Postelsia transect surveys

One of the primary datasets is permanent Postelsia plots---2 m wide band transects from 5 to 10 m in length that run across patches of Postelsia. The number of Postelsia plots varies from 1 - 3 plots per site. First, lets look at a breakdown of the number of sites.

```{r}
#| message: false
#Load files
postelsia_raw <- read_csv("data/raw/postelsia_counts_2024Nov1.csv")
```

```{r}
site_summary <- postelsia_raw %>% 
  select(marine_site_name, marine_common_year, season_name, plot_code, total, plot_area_m2) %>% 
  mutate(density = total/plot_area_m2) %>% 
   group_by(marine_site_name) %>%
  summarize(n_years = n_distinct(marine_common_year),
            n_years_post2004 = n_distinct(marine_common_year[marine_common_year > 2004])
  ) %>% 
  arrange(desc(n_years))

print(site_summary)
```

***Question: Should we set a threshold of the number of years required for sampling such that they can be included as a focal site in analysis? Should we only have a few focal sites, or do we want to share all the data? (i.e. as supplemental to the main sites, which are the ones we will analyze because of the long time series)***

Here are the trends for the 7 sites with \>15 years of data since 2004:

```{r}

postelsia_summary <- postelsia_raw %>% 
  select(marine_site_name, marine_common_year, season_name, plot_code, total, plot_area_m2) %>% 
  mutate(density = total/plot_area_m2) %>% 
  #Filter for only sites with 15+ years of data
  group_by(marine_site_name) %>%
  filter(n_distinct(marine_common_year) > 15) %>%
  ungroup() %>% 
  #Filter for surveys after 2004
  filter(marine_common_year >= 2004) %>% 
  mutate(marine_site_name = factor(marine_site_name),
#         marine_common_year = factor(marine_common_year),
         season_name = factor(season_name), 
         plot_code = factor(plot_code))
  

ggplot(postelsia_summary,
       aes(x=marine_common_year, y=density, color = season_name))+
  # Add semi-transparent red box spanning 2014-2016
  annotate("rect", xmin = 2014, xmax = 2016, ymin = -Inf, ymax = Inf, 
           fill = "red", alpha = 0.2) +
  geom_point()+
#  geom_vline(xintercept = 2015)+
#  stat_smooth(method = 'loess', color = "darkgreen")+
  geom_smooth()+
  
  facet_wrap(facets = "marine_site_name", scales = "free_y")+
  theme_few()
```

Note: the plot above has all seasons mixed into one figure.

***Question: Do we want to choose 1 "canonical" season for the surveys (for each site), or have survey season as a predictor in our modeling?***

Primary survey seasons:

-   Bodega -- Summer

-   Fogarty Creek -- Summer

-   Point Sierra Nevada -- Fall (more years) or Spring (more signal)?

-   Sandhill Bluff -- Spring

-   Scott Creek -- Fall or Spring ?

-   Sea Ranch -- Summer

-   Shelter Cove -- Summer

------------------------------------------------------------------------

#### Dataset 2: Repeat Photos

One potentially valuable dataset that we need to strongly consider whether we include it or not is the repeated photos data. Some of the photos have been manually delineated to determine the % cover of postelsia + mussels. However, not all photos have been done and only at a handful of sites, and there is some funkiness to the data. Here's a coarse breakdown of the sampling coverage:

```{r}
#| message: false
pct_cover_photos <- read_csv("data/raw/pct_cover_photos_raw.csv") %>% 
  mutate(photo = paste(marine_site_name, pan, sep = "_")) 
```

```{r}
print(pct_cover_photos %>% 
        group_by(photo) %>% 
        summarise(n_years = length(unique(year))))
```

Visualize trends from repeat photos split up by site:

```{r}

ggplot(pct_cover_photos, aes(x=year, y=pct_postelsia)) +
    annotate("rect", xmin = 2014, xmax = 2016, ymin = -Inf, ymax = Inf, 
           fill = "red", alpha = 0.2) +
  geom_point()+
  geom_smooth()+
  facet_wrap(facets = "photo", scales = "free_y")+
  theme_few()

```

Visualize repeat photo trends not split by site:

```{r}
ggplot(pct_cover_photos, aes(x=year, y=pct_postelsia)) +
    annotate("rect", xmin = 2014, xmax = 2016, ymin = -Inf, ymax = Inf, 
           fill = "red", alpha = 0.2) +
  geom_point()+
  geom_smooth()+
  theme_few()

ggplot(pct_cover_photos, aes(x=year, y=pct_mussel)) +
    annotate("rect", xmin = 2014, xmax = 2016, ymin = -Inf, ymax = Inf, 
           fill = "red", alpha = 0.2) +
  geom_point()+
  geom_smooth()+
  theme_few()
```

***Question: Do we want to use these data? If so, how confident are we in the extracted values and reproduceability? Would we have to extract percent cover from all photos from the focal sites (and if so, who will do that and when/how)?***

There is some funkiness in these data. For example, some site/pan_ID combinations have 2 rows with different values, and some combinations have 2 rows with the same values.

***Question: If we do not use these data, how will we quantify mussel expansion? Could use biodiversity surveys...if there has been sufficient sampling (?). What about percent cover of mussels in photo quads across the intertidal zone (i.e. across all the targeted assemblages)?***

------------------------------------------------------------------------

#### Dataset 3: Sea Stars

```{r}
#| message: false
stars_raw <- read_csv("data/raw/seastarkat_size_count_zeroes_totals.csv")

stars <- stars_raw %>% 
  #Filter for only postelsia sites
  filter(site_code %in% c("FOG", "SHT", "SEA", "BML", "SCT","PSN")) %>% 
  #Filter for only Pisaster ochraceus
  filter(species_code=="PISOCH") %>% 
  #Filter for surveys after 2004
  filter(marine_common_year >= 2004) %>% 
  #Remove rows with blank star counts
  drop_na(size_bin) %>% 
  filter(size_bin != "NM") %>% 
  #Make columns numeric
  mutate(size_bin = as.numeric(size_bin))%>% 
  #Calculate pisaster biomass using equation from Moritsch and Raimondi (2018) 
  #(https://doi.org/10.1002/ece3.3953)
  mutate(pisaster_biomass_g = total * exp(log(size_bin)*2.34723-5.50262))


stars_summary <- stars %>% 
  group_by(marine_site_name, marine_common_year, marine_common_season) %>% 
  summarise(sum_pisaster_biomass_g = sum(pisaster_biomass_g)) %>% 
  mutate(survey_id = paste(marine_site_name,
                           marine_common_season, 
                           sep = "_"))
```

```{r}

ggplot(stars_summary, aes(x=marine_common_year, y=sum_pisaster_biomass_g))+
  annotate("rect", xmin = 2014, xmax = 2016, ymin = -Inf, ymax = Inf, 
           fill = "red", alpha = 0.2) +
  geom_point()+
  geom_smooth()+
  facet_wrap(facets = "marine_site_name",
             scales = "free_y"
           )+
  theme_few()


```

Note that we do not have star data from Sandhill Bluff.

***Question: Need search area for star quads to standardize pisaster biomass across sites. How do you typically do this?***

------------------------------------------------------------------------

#### **Other data to draw from?** LMK!
